---
title: "Exploring margin of error in ACS drill-downs"
author: Jane Yang
date: 9 August 2023
output: github_document
---

``` {r option_chunks, include = FALSE}
knitr::opts_chunk$set(include = FALSE, message = FALSE)
```

``` {r setup}
#### set up ####
# references
# http://gdfe.co/srvyr/reference/index.html
# https://ldaly.github.io/giveinandblogit/
# https://www.census.gov/programs-surveys/acs/library/handbooks/general.html 

# clear environment and console
rm(list = ls())
cat("\014")

# load libraries
library(tidyverse) # the most useful functions!
library(srvyr) # for easy survey data manipulation
library(rmarkdown) # for GitHub publishing
library(scales) # for better scale labeling
library(extrafont) # custom fonts

# define directories
wd <- if_else(
    str_detect(getwd(), "GitHub"),
    getwd(),
    paste(
        getwd(), 
        "GitHub/acf_analyses/ecd/b_5_explorer", 
        sep = "/"
    )
)
dd <- paste(wd, "data", sep = "/")
od <- paste(wd, "output", sep = "/")

# set design factors
acf_palette <- c(
    "#264A64", # ACF blue
    "#336A90", # medium blue
    "#BCD9ED", # pale blue
    "#BFB0A3", # neutral
    "#DDE2E8", # light cool gray
    "#475260", # medium gray
    "#63BAB0", # teal
    "#F9E585", # yellow
    "#A12854", # magenta
    "#E29F4D" # orange
)

loadfonts(device = "win")

theme_set(theme_minimal(base_family = "Gill Sans MT"))
```

The draft *Birth through Five Population Explorer* is based on the 2017-2021 American Community Survey (ACS) data, downloaded from [IPUMS USA](https://usa.ipums.org/usa/). We joined the ACS data with a state name to FIPS mapping and 2021 state median incomes for a household of four people used by the Department of Health and Human Services (HHS) for the Low Income Home Energy Assistance Program (LIHEAP). 

``` {r load_and_prep}
#### load data ####
# acs 2017-2021 five-year file
d_b5_raw <- read_csv(paste(dd, "acs_2017_2021_rw.csv", sep = "/"))

# state FIPs and median income data
d_states <- read_csv(paste(dd, "state_fips_and_smi4.csv", sep = "/")) %>%
    rename(
        state_code = Code,
        state = State,
        smi4 = `4 Person SMI`,
        statefip = `State FIP`
    )

#### prep data ####
d_b5 <- d_b5_raw 

# lower-case variables names for easier handling
names(d_b5) <- str_to_lower(names(d_b5))

# add in state names
d_b5 <- d_b5 %>% left_join(d_states, by = "statefip")
```

Using the [successive difference replication variance methodology](https://www2.census.gov/programs-surveys/acs/tech_docs/accuracy/2021_ACS_Accuracy_Document_Worked_Examples.pdf), we can calculate margins of errors for population estimates for specific cross-sections of children. 

``` {r calculate_margin_of_errors, include = TRUE}
#### examine margin of errors in population estimates ####
# create survey object
d_b5_svy <- as_survey(d_b5, weight = perwt, repweights = matches("repwtp[0-9]+"),
    type = "JK1", scale = 4/ 80 , rscales = rep(1, 80 ), mse = TRUE)

# calculate margin of error when aggregating by puma, age year, and
# household income <= 100% of poverty level
s_b5 <- d_b5_svy %>%
    mutate(household_income_less_than_100pct_poverty = poverty <= 100) %>%
    group_by(
        state, statefip, puma,
        age,
        household_income_less_than_100pct_poverty,
    ) %>%
    summarize(
        # estimate population with standard error
        population_estimate = survey_total(vartype = "se"),
        # calculate sample size for each aggregation
        acs_sample_size = n(),
    ) %>%
    ungroup()

s_b5 <- s_b5 %>% 
    mutate(
        # multiply SE by 1.645 to get MOE at 90% conf level (Census standard)
        population_estimate_moe = population_estimate_se * 1.645,
        population_estimate_moe_pct_of_estimate = population_estimate_moe/population_estimate,
    )
```

For instance, we can examine the margin of errors (MOE) as a percent of the population estimate for groupings of [public use microdata areas (PUMAs)](https://www.census.gov/programs-surveys/geography/guidance/geo-areas/pumas.html), single-year ages, and whether a child lives in a household with <100% the Federal poverty line. There are `r comma(nrow(s_b5))` such groupings. The following histogram illustrates the distribution.


``` {r plot_moe_distribution, include = TRUE}
# visualize distribution of MOE as pct of estimate
g_moe_pct <- ggplot(s_b5) +
    geom_histogram(
        aes(x = population_estimate_moe_pct_of_estimate),
        binwidth = 0.01,
        fill = acf_palette[2]
    ) +
    labs(
        title = "Distribution of margins of errors (MOE) as percent of population estimate",
        subtitle = "by groupings of PUMA, single-year age, and\nhousehold income greater or less than 100% of Federal poverty line",
        x = "MOE/population estimate",
        y = "number of groupings"
    ) +
    scale_x_continuous(labels = label_percent(accuracy = 1))

g_moe_pct
```

``` {r illustrate_moe_as_percent_of_estimate}
# calculate percent of puma/age/100pct_poverty aggregates with
# MOE > 50% , > 100% of population estimate
pct_moe_over_estimate <- s_b5 %>%
    mutate(
        moe_over_50pct_estimate = population_estimate_moe_pct_of_estimate > 0.5,
        moe_over_100pct_estimate = population_estimate_moe_pct_of_estimate > 1,
    ) %>%
    summarize(
        over_50pct = mean(moe_over_50pct_estimate),
        over_100pct = mean(moe_over_100pct_estimate),
    )
```

Of the combinations of PUMA, age by year, and indicator of whether a child lives in a household with <100% the Federal poverty line, `r percent(pct_moe_over_estimate$over_100pct, accuracy = 1)` have an MOE > 100% of the population estimate and `r percent(pct_moe_over_estimate$over_50pct, accuracy = 1)` have an MOE > 50% of the population estimate.

The draft *B-5 Explorer* allows for significantly more filters to be applied to a population grouping, which results in even smaller sample sizes behind each population estimate. The MOE need to be further examined for cases with more granular filters and the filters themselves should be reconsidered for the highest priority ones that would be useful for lead agencies serving children.

``` {r save_data}
#### save data ####
#write_csv(s_b5, paste(od, "b5_2021_moe.csv", sep = "/"))
#save(d_b5, s_b5, file = paste(dd, "b5_2021.Rdata", sep = "/"))
```